Detailed structure of negatively charged membranes of phosphatidylserine and phosphatidylcholine at concentrations of calcium, sodium and potassium salts from molecular dynamics simulations with electronic polarization

### 2019-05-21 Manuscript/manuscript.tex
Edited monovalent ions section

\subsection{Measurements of acyl chain order parameters}
A R-type Proton Detected Local Field (R-PDLF) experiment was performed to determine C--H bond order
parameters
\begin{equation}\label{OPequation}
  S_{\rm{CH}}=\frac{1}{2}\langle3\cos^2\theta-1\rangle,
\end{equation}
  where $\theta$ denotes the angle of the C--H bond
with the bilayer normal and the angular brackets define a time average on a time scale of approximately 1 microsecond.
The experiment was done using a Bruker Avance III 400 spectrometer operating at a 1H Larmor frequency of 400.03 MHz
equipped with a standard 4~mm CP-MAS HXY probe. The set up of the R-PDLF experiment was the following
(using the notation from figures 1c and 2c in the original publication describing the R-PDLF experiment \cite{dvinskikh04}).
The magic angle spinning (MAS) frequency used was 5.15 kHz. The recoupling pulses in the R18 blocks had
therefore a length of $10.79 \, \mu \mathrm{s}$, which correspond to a nutation frequency of 46.35~kHz. Increments of $18 \times 2$ recoupling
pulses where used giving a spectral width of 2.6~kHz in the indirect dimension and a total number of 32 points in
the indirect dimension were recorded. These settings enable to record dipolar slices with a C--H bond order parameter
resolution of $\pm 0.01$. The C--H bond order parameter determined from a given dipolar
splitting, $\Delta \nu$ (see e.g. Fig. \ref{R-PDLFslices} in SI)
is equal to $\Delta \nu / (0.315 \times 21.5 \, \mathrm{kHz})$, where 0.315 is
the scaling factor of the R18 recoupling sequence and 21.5 kHz is the maximum $^1$H-$^{13}$C dipolar coupling for a C--H bond.
The refocused-INEPT transfer \cite{morris79,burum80} %[G. Morris and R. Freeman, J. Am. Chem. Soc. 101, 760 (1979);
was used for transferring the polarization from $^1$H nuclei to the covalently bond $^{13}$C nuclei with the delays $\tau_1 = 1.94$~ms and $\tau_2 = 0.97$~ms
(set as multiples of the MAS rotation period). The pulses for the refocused-INEPT sequence had a nutation frequency equal to 63.45~kHz.
For acquiring the spectra, a total number of 1024 transients were recorded for each point in the indirect dimension, using an acquisition
time of 0.1~s under SPINAL64 $^1$H decoupling \cite{fung00} with a nutation frequency of 50~kHz, and with dwell time giving a spectral
width of 200 ppm for the $^{13}$C chemical shift dimension. The set of free induction decays recorded were then processed as
described e.g. in Ref.~\citenum{ferreira08}. The dipolar splittings in the crowded spectral region
between 29 and 30 ppm were assigned based on the previous assignments for POPC \cite{ferreira13} 
and the POPS order parameters calculated from simulations in this work.

@@

The X-ray scattering form factors and C-H bond order parameters from NMR experiments are good measures to validate
the lipid bilayer structure in MD simulations against experiments because they
are related to the bilayer dimensions (area per molecule and thickness) and
conformational fluctuations of individual lipids, respectively, and 
can be directly calculated from MD simulations~\cite{ollila16}.
The X-ray scattering form factors and C-H bond order parameters in
the headgroup and glycerol backbone region of a POPS lipid bilayer are already available in the literature \cite{kucerka14,NMRlipidsIV}.
Here, we measure also the C--H bond order parameters of the acyl chain region using natural abundance $^{13}$C NMR.
We first determined the order parameters of each peak corresponding to different C--H bonds detected
in the first dimension (Fig. \ref{R-PDLF}) from the dipolar slices in the indirect dimension (Fig. \ref{R-PDLFslices}).
Because the exact assignment of these peaks was not possible, we assigned the
acyl chain order parameters to mimic the profile predicted by the MD simulations (Fig. \ref{simVSexpNOions_POPS}).
\todo{Tiago has to check and update this and figures in SI.}

@@

The Lipid17/Dang simulation gives larger acyl chain order parameters and smaller area per molecule than the
experimental values (Fig. \ref{simVSexpNOions_POPS} and table \ref{tab:apls}),
indicating that this simulation predicts a too compact POPS lipid bilayer. This is also in line with
the discrepancy observed in the X-ray form factor between the simulation and experiment (Fig.~\ref{simVSexpNOions_POPS}).
In previous study \cite{NMRlipidsIV}, the Lipid17 force field with {\AA}qvist ions~\cite{aqvist90}
gave a larger area per molecule (57~\AA$^2$) for POPS being closer to the experimental value,
which is probably due to the weaker counterion binding affinity of {\AA}qvist ions to phospholipids. \cite{catte16, NMRlipidsIV}
However,  {\AA}qvist ions produce known artifacts, 
e.g. they exhibit artificial aggregation of ions at larger salt concentrations \cite{kohagen16, chen07, NMRlipidsIV}.
Therefore, the Dang ions with more realisitic bulk properties are used here to analyze the effect of ECC to the properties of a POPS lipid bilayer. 
The ECC-POPS simulation gives a good agreement with the X-ray scattering form factors and experimental
acyl chain order parameters (see Fig.~\ref{simVSexpNOions_POPS}) indicating that the bilayer dimensions and acyl chain
conformations are well described by the force field. The area per lipid from ECC-POPS simulation is slightly
smaller than the value reported from SDP model \cite{kucerka14} (table~\ref{tab:apls}), but the values
agree within their error estimates.

@@

Despite the success in simulating bilayer dimensions and acyl chain conformations,
lipid force fields have typically problems in capturing the correct glycerol backbone and
headgroup order parameters and conformations \cite{botan15,ollila16,NMRlipidsIV}.
This is the case also for Lipid17 POPS simulations, where the headgroup and glycerol backbone
order parameters are quite far from experimental values with both Dang (Fig. \ref{simVSexpNOions_POPS})
and {\AA}qvist ions. \cite{NMRlipidsIV} The values from ECC-POPS simulation are closer,
but not in full agreement with experiments (Fig.~\ref{simVSexpNOions_POPS}).
Especially, the experimentally observed more negative values of the $\beta$-carbon
and the larger forking (difference of order parameters between hydrogens attached to the
same carbon \cite{ollila16}) of the $\alpha$-carbon in POPS than in POPC \cite{NMRlipidsIV}
are not reproduced by neither of the models here. 
These differences are previously related
to a more rigid structure of PS headgroup \cite{browning80,buldt81} and 
the stiffness of the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ dihedral \cite{NMRlipidsIV}
plotted in Fig.~\ref{fig:dihedral}. 
Interestingely, this dihedral angle is more flexible in ECC-POPS compared to Lipid17 
suggesting that its rigidity is overestimated in Lipid17 due to the lack of electronic polarizability,
while the same dihedral parameters yield too small forking in ECC-POPS.

Overall,
the head group and acyl chain order parameters within ECC-lipids
are in a good agreement with the experimental values 
as shown in Fig. \ref{simVSexpNOions_POPS}. 
The acyl chain order parameters in particular agree with the experimental data within error bars.
The order parameters of the head groups are at an accuracy comparable to 
other currently available classical models of lipids \citep{botan15, catte16, Pluhackova2016, nmrlipids_proj4}. 
In particular, ECC-POPS is closer to the experimental order parameters 
than Lipid17 with either Dang \cite{smith94,chang1999,dang2006} 
or {\AA}qvist ions \cite{aqvist90}. 
Although the ECC-POPS model reproduces the lipid bilayer dimensions and
acyl chain conformations in a good agreement with experiments, 
there is a room for improvements especially in the glycerol backbone and headgroup region,
e.g. in the parameters of the associated dihedral angles. 
\cite{botan15, ollila16, Pluhackova2016, NMRlipidsIV}

@@

\begin{figure*}[!tbp] 
  \centering 
  \includegraphics[height=\figheightsmall]{../img/ecc_pops/density_profiles_na-k-counterions_wat_phos_compar_5PC-1PS_ecclipids-lipid17.pdf}
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_nacl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_kcl.pdf} 
  \includegraphics[height=\figheightsmall]{../img/POPCstructure.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PC-PS_mix_PC} 
    \textbf{Left:} Number density profiles of \ce{K^{+}} and \ce{Na^{+}} counterions along the membrane normal axis
    in ECC-lipids (solid lines) and Lipid17/Dang (dashed lines) simulations of POPC:POPS (5:1) bilayers.  
    The density profiles of phosphate groups and water are divided by 4 and 100, respectively.  
    \textbf{Middle:} The POPC head group order parameters and the P--N vector angle
    with respect to the membrane normal as a function of POPS content in a bilayer
    from ECC-lipids and Lipid17/Dang simulations with \ce{Na^+} (left) and \ce{K^+} (right) counterions.
    Experimental order parameter values are from Ref. \citenum{scherer87}
    and the signs from Ref. \citenum{ferreira16} 
    (only \ce{Na^+} counterions, but shown also in the right plots for \ce{K+}).
    \textbf{Right: Chemical structure and labelling of carbon segments of POPC. }
  }
\todo{J: Will manually improve this figure with panels A, B, C and D 
and will erase the duplicate y-axis description 
and will add ticks to the right side of the plotting area. }
\end{figure*}

First, we monitor the response of POPC headgroup order parameters to the increasing amount of POPS (Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix}).
According to the electrometer concept, the head group order parameters of PC lipids increase
with the increasing amount of negatively charged PS lipids in experiments, because the headgroup P--N dipole
tilts more parallel to the membrane interface \cite{seelig87,scherer87}.
The trend is observed with both force fields in simulations with \ce{K+} counterions and
in ECC-POPS simulation with \ce{Na+} counterions, but not in Lipid17/Dang ion simulations with \ce{Na+} counterions,
where the order parameters are unchanged or decrease and the P--N vector tilts more perpendicular to the membrane interface.
The result can be explained by the higher binding affinity of sodium in the
Lipid17/Dang ion simulations (Fig~\ref{fig:delta_ordPar_NaCl_PC-PS_mix_PC}), which
compensates the negative charge of PS lipids at the interface.
In contrast, ECC-lipids represent this trend correctly
with only slightly underestimated slopes of the order parameters 
when \ce{Na+} counterions are used (Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix}).

Second, we monitor the response of PC headgroup order parameters to the additional
amount of sodium and potassium in POPC:POPS (5:1) mixtures.
In experiments, the addition of K$^+$ ions led to a very modest decrease of the order parameters,
while the decrease was more pronounced with Li$^+$ ions and was the strongest with divalent Ca$^{2+}$ and Mg$^{2+}$ ions \cite{roux90}.
According to the electrometer concept, the decrease of PC lipid head group order parameters
correlates with the amount of bound cations \cite{akutsu81,altenbach84,seelig87,roux90,catte16,NMRlipidsIV},
thus the binding affinity of the ions to POPC:POPS (5:1) mixtures increase in order K$^{+}$ $<$ Li$^{+}$  $<$ Mg$^{2+}$  $<$ Ca$^{2+}$.
The response of PC headgroup order parameters to the increasing amount of KCl is close to experiments
and the change in P--N vector is negligible in both force fields (Fig. \ref{fig:delta_ordPar_monoval_PCPS}), athough the change of the $\alpha$-carbon
order parameter may be slightly overestimated by the Lipid17/Dang simulations. 
This could be explained by the slightly deeper penetration of potassium into the bilayer (Fig. \ref{fig:delta_ordPar_NaCl_PC-PS_mix_PC}).

Because experimental data with the additional amount of sodium in POPC:POPS (5:1) mixture was not available,
we compared the simulation results to the experimental data with the additional amounts of
lithium and potassium ions, which should give the lower and upper bounds to the sodium binding
affinity (Fig. \ref{fig:delta_ordPar_monoval_PCPS}). The PC headgroup response to the additional
amount of sodium is clearly overestimated by the Lipid17/Dang simulation and
the response in ECC-lipids simulations is probably too close to the experimental results for lithium.
% The following conclusion is already in the previous sentence, so I commented it out
%In conclusion, the results suggest that the sodium binding affinity (Fig. \ref{fig:POPS-counterions-dens}) and
%the tilting of P--N vector more perpendicular to the membrane surface (Fig. \ref{fig:delta_ordPar_NaCl_PCPS})
%are clearly overestimated in Lipid17/Dang simulations and slightly overestimated in ECC-POPS simulations,
%while the binding affinity of potassium is better described by both force fields.

@@

The changes of POPS headgroup order parameters are almost negligible in experiments with increasing POPC fraction in bilayer
or monovalent ion concentration in solvent
(Figs. \ref{fig:delta_ordPar_monoval_PCPS}, \ref{fig:delta_ordPar_monoval_PCPS} and \ref{fig:delta_ordPar_NaCl_PC-PS_mix}).
%while calcium leads to relatively large changes with low concentrations followed by a rapid saturation below 100mM (Fig. \ref{fig:delta_ordPar_CaCl_PCPS}) \cite{scherer87,roux90}.
%The interaction with \ce{K^+}, which binds very weakly to both neutral and negatively charged membranes, 
%renders a qualitatively different response of the order parameter $S^\beta$ in POPS 
%in the mixed negatively charged bilayers compared to the neutral bilayers. 
%While the order parameter $S^\beta$ increases for both \ce{Na^+} and \ce{Ca^{2+}},
%it decreases in the presence of \ce{K^+}
%(\ce{KCl}:  Fig.~\ref{fig:delta_ordPar_KCl_PCPS}, 
% \ce{NaCl}: Fig.~\ref{fig:delta_ordPar_NaCl_PCPS}, and
%\ce{CaCl2}: Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}). 
In Lipid17/Dang simulations, the changes of POPS headgroup order parameters with increasing amount of POPC
are overestimated, which is also related to the changes in the orientations of the P--N and C$_{\beta}$--C$_{\gamma}$
vectors (Fig. \ref{fig:delta_ordPar_NaCl_PC-PS_mix}). The overestimation of the order parameter changes
upon increasing amount of POPC in a bilayer 
was also observed in previous simulations with other force fields \cite{NMRlipidsIV}, but not in
ECC-POPS simulation, where also the changes in the PS headgroup orientation are more modest (Fig. \ref{fig:delta_ordPar_NaCl_PC-PS_mix}).
The POPS headgroup orientation and order parameters are more sensitive also to the additional amount
of sodium and potassium in Lipid17/Dang simulations, while dependence is weaker and closer to experiments
in ECC-POPS simulations (Figs. \ref{fig:delta_ordPar_monoval_PCPS} and \ref{fig:delta_ordPar_monoval_PCPS}).
These structural changes can relate to variations in the interactions between PC and PS head groups 
and also to the changing amount of counterions and PS in the composition. 
%In contrast, ECC-lipids with ECC-ions capture the different response 
%of the order parameters $S^{\beta}$, $S^{\alpha _1}$ and $S^{\alpha _2}$ in POPS 
%to various concentrations of \ce{NaCl} and \ce{KCl} in a good agreement with the experiments
%(\ce{KCl}:  Fig.~\ref{fig:delta_ordPar_KCl_PCPS}, 
% \ce{NaCl}: Fig.~\ref{fig:delta_ordPar_NaCl_PCPS}). 
%Such a dramatic improvement in the structural parameters 
%demonstrates that including electronic polarization in phospholipids is crucial 
%even for interactions with soft cations like \ce{K^+}.

>>>>

\subsection{Measurements of acyl chain order parameters}
A R-type Proton Detected Local Field (R-PDLF) experiment was performed to determine C--H bond order
parameters
\begin{equation}\label{OPequation}
  S_{\rm{CH}}=\frac{1}{2}\langle3\cos^2\theta-1\rangle,
\end{equation}
  where $\theta$ denotes the angle of the C--H bond
with the bilayer normal and the angular brackets define a time average on a time scale of approximately 1 microsecond.
The experiment was done using a Bruker Avance III 400 spectrometer operating at a 1H Larmor frequency of 400.03 MHz
equipped with a standard 4~mm CP-MAS HXY probe. The set up of the R-PDLF experiment was the following
(using the notation from figures 1c and 2c in the original publication describing the R-PDLF experiment \cite{dvinskikh04}).
The magic angle spinning (MAS) frequency used was 5.15 kHz. The recoupling pulses in the R18 blocks had
therefore a length of $10.79 \, \mu \mathrm{s}$, which correspond to a nutation frequency of 46.35~kHz. Increments of $18 \times 2$ recoupling
pulses where used giving a spectral width of 2.6~kHz in the indirect dimension and a total number of 32 points in
the indirect dimension were recorded. These settings enable to record dipolar slices with a C--H bond order parameter
resolution of $\pm 0.01$. The C--H bond order parameter determined from a given dipolar
splitting, $\Delta \nu$ (see e.g. Fig.~\ref{R-PDLFslices} in SI)
is equal to $\Delta \nu / (0.315 \times 21.5 \, \mathrm{kHz})$, where 0.315 is
the scaling factor of the R18 recoupling sequence and 21.5 kHz is the maximum $^1$H-$^{13}$C dipolar coupling for a C--H bond.
The refocused-INEPT transfer \cite{morris79,burum80} %[G. Morris and R. Freeman, J. Am. Chem. Soc. 101, 760 (1979);
was used for transferring the polarization from $^1$H nuclei to the covalently bond $^{13}$C nuclei with the delays $\tau_1 = 1.94$~ms and $\tau_2 = 0.97$~ms
(set as multiples of the MAS rotation period). The pulses for the refocused-INEPT sequence had a nutation frequency equal to 63.45~kHz.
For acquiring the spectra, a total number of 1024 transients were recorded for each point in the indirect dimension, using an acquisition
time of 0.1~s under SPINAL64 $^1$H decoupling \cite{fung00} with a nutation frequency of 50~kHz, and with dwell time giving a spectral
width of 200 ppm for the $^{13}$C chemical shift dimension. The set of free induction decays recorded were then processed as
described e.g. in Ref.~\citenum{ferreira08}. The dipolar splittings in the crowded spectral region
between 29 and 30 ppm were assigned based on the previous assignments for POPC \cite{ferreira13} 
and the POPS order parameters calculated from simulations in this work.

@@

The X-ray scattering form factors and C-H bond order parameters from NMR experiments are good measures to validate
the lipid bilayer structure in MD simulations against experiments because they
are related to the bilayer dimensions (area per molecule and thickness) and
conformational fluctuations of individual lipids, respectively, and 
can be directly calculated from MD simulations~\cite{ollila16}.
The X-ray scattering form factors and C-H bond order parameters in
the headgroup and glycerol backbone region of a POPS lipid bilayer are already available in the literature \cite{kucerka14,NMRlipidsIV}.
Here, we measure also the C--H bond order parameters of the acyl chain region using natural abundance $^{13}$C NMR.
We first determined the order parameters of each peak corresponding to different C--H bonds detected
in the first dimension (Fig.~\ref{R-PDLF}) from the dipolar slices in the indirect dimension (Fig.~\ref{R-PDLFslices}).
Because the exact assignment of these peaks was not possible, we assigned the
acyl chain order parameters to mimic the profile predicted by the MD simulations (Fig.~\ref{simVSexpNOions_POPS}).
\todo{Tiago has to check and update this and figures in SI.}

@@

The Lipid17/Dang simulation gives larger acyl chain order parameters and smaller area per molecule than the
experimental values (Fig.~\ref{simVSexpNOions_POPS} and table~\ref{tab:apls}),
indicating that this simulation predicts a too compact POPS lipid bilayer. This is also in line with
the discrepancy observed in the X-ray form factor between the simulation and experiment (Fig.~\ref{simVSexpNOions_POPS}).
In previous study \cite{NMRlipidsIV}, the Lipid17 force field with {\AA}qvist ions~\cite{aqvist90}
gave a larger area per molecule (57~\AA$^2$) for POPS being closer to the experimental value,
which is probably due to the weaker counterion binding affinity of {\AA}qvist ions to phospholipids. \cite{catte16, NMRlipidsIV}
However,  {\AA}qvist ions produce known artifacts, 
e.g. they exhibit artificial aggregation of ions at larger salt concentrations \cite{kohagen16, chen07, NMRlipidsIV}.
Therefore, the Dang ions with more realisitic bulk properties are used here to analyze the effect of ECC to the properties of a POPS lipid bilayer. 
The ECC-POPS simulation gives a good agreement with the X-ray scattering form factors and experimental
acyl chain order parameters (see Fig.~\ref{simVSexpNOions_POPS}) indicating that the bilayer dimensions and acyl chain
conformations are well described by the force field. The area per lipid from ECC-POPS simulation is slightly
smaller than the value reported from SDP model \cite{kucerka14} (table~\ref{tab:apls}), but the values
agree within their error estimates.

@@

Despite the success in simulating bilayer dimensions and acyl chain conformations,
lipid force fields have typically problems in capturing the correct glycerol backbone and
headgroup order parameters and conformations \cite{botan15,ollila16,NMRlipidsIV}.
This is the case also for Lipid17 POPS simulations, where the headgroup and glycerol backbone
order parameters are quite far from experimental values with both Dang (Fig.~\ref{simVSexpNOions_POPS})
and {\AA}qvist ions. \cite{NMRlipidsIV} The values from ECC-POPS simulation are closer,
but not in full agreement with experiments (Fig.~\ref{simVSexpNOions_POPS}).
Especially, the experimentally observed more negative values of the $\beta$-carbon
and the larger forking (difference of order parameters between hydrogens attached to the
same carbon \cite{ollila16}) of the $\alpha$-carbon in POPS than in POPC \cite{NMRlipidsIV}
are not reproduced by neither of the models here. 
These differences are previously related
to a more rigid structure of PS headgroup \cite{browning80,buldt81} and 
the stiffness of the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$ dihedral \cite{NMRlipidsIV}
plotted in Fig.~\ref{fig:dihedral}. 
Interestingely, this dihedral angle is more flexible in ECC-POPS compared to Lipid17 
suggesting that its rigidity is overestimated in Lipid17 due to the lack of electronic polarizability,
while the same dihedral parameters yield too small forking in ECC-POPS.

Overall,
the head group and acyl chain order parameters within ECC-lipids
are in a good agreement with the experimental values 
as shown in Fig.~\ref{simVSexpNOions_POPS}. 
The acyl chain order parameters in particular agree with the experimental data within error bars.
The order parameters of the head groups are at an accuracy comparable to 
other currently available classical models of lipids \citep{botan15, catte16, Pluhackova2016, nmrlipids_proj4}. 
In particular, ECC-POPS is closer to the experimental order parameters 
than Lipid17 with either Dang \cite{smith94,chang1999,dang2006} 
or {\AA}qvist ions \cite{aqvist90}. 
Although the ECC-POPS model reproduces the lipid bilayer dimensions and
acyl chain conformations in a good agreement with experiments, 
there is a room for improvements especially in the glycerol backbone and headgroup region,
e.g. in the parameters of the associated dihedral angles. 
\cite{botan15, ollila16, Pluhackova2016, NMRlipidsIV}

@@

\begin{figure*}[!tbp] 
  \centering 
  \includegraphics[height=\figheightsmall]{../img/ecc_pops/density_profiles_na-k-counterions_wat_phos_compar_5PC-1PS_ecclipids-lipid17.pdf}
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_nacl.pdf} 
  \includegraphics[height=\figheightsmall]{../Fig/order_parameters_changes_A-B_PC-PS_mix_POPC_kcl.pdf} 
  \includegraphics[height=\figheightsmall]{../img/POPCstructure.pdf} 
  \caption{\label{fig:delta_ordPar_NaCl_PC-PS_mix_PC} 
    \textbf{Left:} Number density profiles of \ce{K^{+}} and \ce{Na^{+}} counterions along the membrane normal axis
    in ECC-lipids (solid lines) and Lipid17/Dang (dashed lines) simulations of POPC:POPS (5:1) bilayers.  
    The density profiles of phosphate groups and water are divided by 4 and 100, respectively.  
    \textbf{Middle:} The POPC head group order parameters and the P--N vector angle
    with respect to the membrane normal as a function of POPS content in a bilayer
    from ECC-lipids and Lipid17/Dang simulations with \ce{Na^+} (left) and \ce{K^+} (right) counterions.
    Experimental order parameter values are from Ref. \citenum{scherer87}
    and the signs from Ref. \citenum{ferreira16} 
    (only \ce{Na^+} counterions, but shown also in the right plots for \ce{K+}).
    \textbf{Right:} Chemical structure and labelling of carbon segments of POPC. 
  }
\todo{J: Will manually improve this figure with panels A, B, C and D 
and will erase the duplicate y-axis description 
and will add ticks to the right side of the plotting area. }
\end{figure*}

First, we monitor the response of POPC headgroup order parameters to the increasing amount of POPS (Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix_PC}).
According to the electrometer concept, the head group order parameters of PC lipids increase
with the increasing amount of negatively charged PS lipids in experiments, because the headgroup P--N dipole
tilts more parallel to the membrane interface \cite{seelig87,scherer87}.
The trend is observed with both force fields in simulations with \ce{K+} counterions and
in ECC-lipids simulation with \ce{Na+} counterions. 
In contrast, the order parameters are unchanged or decrease 
and the P--N vector tilts more perpendicular to the membrane interface.
in Lipid17/Dang ion simulations with \ce{Na+} counterions.
The result can be explained by the higher binding affinity of sodium in the
Lipid17/Dang ion simulations (Fig~\ref{fig:delta_ordPar_NaCl_PC-PS_mix_PC}), which
compensates the negative charge of PS lipids at the interface.
This is corrected in ECC-lipids, 
which only slightly underestimate the slopes of the order parameters 
when \ce{Na+} counterions are used (Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix_PC}).

Second, we monitor the response of PC headgroup order parameters to the additional
amount of \ce{NaCl} and \ce{KCl} in POPC:POPS (5:1) mixtures.
In experiments, the addition of K$^+$ ions led to a very modest decrease of the order parameters,
while the decrease was more pronounced with Li$^+$ ions and was the strongest with divalent Ca$^{2+}$ and Mg$^{2+}$ ions \cite{roux90}.
According to the electrometer concept, the decrease of PC lipid head group order parameters
correlates with the amount of bound cations \cite{akutsu81,altenbach84,seelig87,roux90,catte16,NMRlipidsIV},
thus the binding affinity of the ions to POPC:POPS (5:1) mixtures increase in order K$^{+}$ $<$ Li$^{+}$  $<$ Mg$^{2+}$  $<$ Ca$^{2+}$.
The response of PC headgroup order parameters to the increasing amount of KCl is close to experiments
and the change in P--N vector is negligible in both force fields (Fig.~\ref{fig:delta_ordPar_monoval_PCPS}), athough the change of the $\alpha$-carbon
order parameter may be slightly overestimated by the Lipid17/Dang simulations. 
This could be explained by the deeper penetration of \ce{K+} into the bilayer (Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix_PC}).

Because experimental data with the additional amount of \ce{Na+} in POPC:POPS (5:1) mixture was not available,
we compared the simulation results to the experimental data with the additional amounts of
\ce{Li+} and \ce{K+} ions, which should give the lower and the upper bounds to the sodium binding
affinity (Fig.~\ref{fig:delta_ordPar_monoval_PCPS}). The PC headgroup response to the additional
amount of sodium is clearly overestimated by the Lipid17/Dang simulation, and
the response in ECC-lipids simulations is probably too close to the experimental results for lithium.
% The following conclusion is already in the previous sentence, so I commented it out
%In conclusion, the results suggest that the sodium binding affinity (Fig. \ref{fig:POPS-counterions-dens}) and
%the tilting of P--N vector more perpendicular to the membrane surface (Fig. \ref{fig:delta_ordPar_NaCl_PCPS})
%are clearly overestimated in Lipid17/Dang simulations and slightly overestimated in ECC-lipids simulations,
%while the binding affinity of potassium is better described by both force fields.

@@

The changes of POPS headgroup order parameters are almost negligible in experiments with increasing POPC fraction in bilayer
or monovalent ion concentration in solvent
(Figs.~\ref{fig:delta_ordPar_monoval_PCPS},~\ref{fig:delta_ordPar_monoval_PCPS} and~\ref{fig:delta_ordPar_NaCl_PC-PS_mix}).
%while calcium leads to relatively large changes with low concentrations followed by a rapid saturation below 100mM (Fig. \ref{fig:delta_ordPar_CaCl_PCPS}) \cite{scherer87,roux90}.
%The interaction with \ce{K^+}, which binds very weakly to both neutral and negatively charged membranes, 
%renders a qualitatively different response of the order parameter $S^\beta$ in POPS 
%in the mixed negatively charged bilayers compared to the neutral bilayers. 
%While the order parameter $S^\beta$ increases for both \ce{Na^+} and \ce{Ca^{2+}},
%it decreases in the presence of \ce{K^+}
%(\ce{KCl}:  Fig.~\ref{fig:delta_ordPar_KCl_PCPS}, 
% \ce{NaCl}: Fig.~\ref{fig:delta_ordPar_NaCl_PCPS}, and
%\ce{CaCl2}: Fig.~\ref{fig:delta_ordPar_CaCl_PCPS}). 
In Lipid17/Dang simulations, the changes of POPS headgroup order parameters with increasing amount of POPC
are overestimated, which is also related to the changes in the orientations of the P--N and C$_{\beta}$--C$_{\gamma}$
vectors (Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix_PC}). 
%These structural changes can relate to variations in the interactions between PC and PS head groups 
%and also to the changing amount of counterions and PS in the composition. 
Also, the PS headgroup order parameters from simulations with Lipid17/Dang model
are too sensitive to an additional amount of \ce{KCl} and \ce{NaCl} 
(Fig.~\ref{fig:delta_ordPar_monoval_PCPS}).
The overestimation of the order parameter changes upon increasing amount of POPC in a bilayer 
was also observed in previous simulations with other force fields \cite{NMRlipidsIV}, 
but not in ECC-lipids simulation, 
where the changes in the PS headgroup orientation are more modest (Fig.~\ref{fig:delta_ordPar_NaCl_PC-PS_mix_PC}).
In particular, 
the dependence of the PS headgroup order parameters from ECC-lipids simulations
to increasing amount of POPC, \ce{KCl} or \ce{NaCl}
is weaker and closer to experiments 
(Fig.~\ref{fig:delta_ordPar_monoval_PCPS}).
%In contrast, ECC-lipids with ECC-ions capture the different response 
%of the order parameters $S^{\beta}$, $S^{\alpha _1}$ and $S^{\alpha _2}$ in POPS 
%to various concentrations of \ce{NaCl} and \ce{KCl} in a good agreement with the experiments
%(\ce{KCl}:  Fig.~\ref{fig:delta_ordPar_KCl_PCPS}, 
% \ce{NaCl}: Fig.~\ref{fig:delta_ordPar_NaCl_PCPS}). 
%Such a dramatic improvement in the structural parameters 
%demonstrates that including electronic polarization in phospholipids is crucial 
%even for interactions with soft cations like \ce{K^+}.

% This§: 
% adsorption and affinities of Na and K to PC:PS memb
% from densities and rel.surf.excess
% 
%in  Fig.~\ref{fig:POPS-counterions-dens} (only counterions)
%and Fig.~\ref{fig:nacl-dens_PCPS}        (additional salt concentrations), 
%
The influence of negatively charged PS on the counterion binding affinity can be estimated 
%by calculating the amounts of phospholipids per bound \ce{Ca^{2+}} cation, $\xi _{Ca^{2+}}$,
by comparing the relative surface excesses with respect to water, $\Gamma ^{w} _{\rm ion}$,
between POPC:POPS (5:1) mixture and pure POPC \cite{melcr18}.
%Threshold for counting bound \ce{Ca^{2+}} cations to a membrane
%was set to $0.3\,\mathrm{nm}$ from oxygen atoms in lipids as in Ref.~\citenum{melcr18}.  
%Such a quantity compares the adsorption of ions to the adsorption of water molecules at an interface 
%without the necessity of defining a Gibbs dividing surface. \citep{melcr18, chattorajBOOK}
The value of $\Gamma ^{w} _{\rm Na}=-0.11 \pm 0.01 \mathrm{nm}^{-2}$ for 1~M sodium in pure ECC-POPC
simulation was reported in the previous work \cite{melcr18}.
The presence of $\sim$17\% POPS in a POPC bilayer increases the
%relative surface excess of sodium with respect to water from negative
value to $\Gamma ^{w} _{\rm Na}=0.092 \pm 0.005 \mathrm{nm}^{-2}$ 
for 0.621 M sodium, while value for potassium remains negative,
$\Gamma ^{w} _{\rm K}=-0.123 \pm 0.005 \mathrm{nm}^{-2}$ in POPC:POPS (5:1) mixture (Fig.~\ref{fig:nacl-dens_PCPS}).
As discussed above, the sodium binding to POPC:POPS (5:1) mixture may be slightly overestimated also in
ECC-lipids simulations. The response of headgroup order parameters to the addition of NaCl are
slightly overestimated also in ECC-POPC simulations (Fig 3 in Ref. \citenum{melcr18}),
suggesting that similar small overestimation of sodium binding may also occur to a pure POPC bilayer.
Therefore, the overestimation may arise from the sodium interactions with PC headgroup and
increase in binding affinity due to PS lipids would not be affected.
%Interestingly enough, \ce{K^+} maintains negative values of $\Gamma^{w}_{K}$ even for the negatively charged PC:PS~(5:1) bilayer.
%This is also evident from the density plots in  Fig.~\ref{fig:nacl-dens_PCPS}
%showing that the density profile of \ce{K^+} along the membrane normal 
%decays clearly faster towards the membrane centre compared to the density of water. 
%In contrast, the density profile of \ce{Na^+} has a shoulder 
%at the region of the phosphate groups,
%which demonstrates that sodium cations adsorb weakly to the membrane interface compared to the aquaeous solution
%by yielding a small positive value of $\Gamma^{w}_{Na}$.
